---
title: "NBA Statistics Project"
format: html
---

## Problem Statements

1.	What is the best shooting position for each player?
2.	At what range each player is most likely to score a shot?
3.	Who of these players is the best defender?
4.	On whom of these players would you put the best defender?
5.	Do the efficiency of a shooter and the player defending him correlate?
6.	How are taken and missed shots distributed spatially on the court?

## Description of the Data

This dataset is based on shooting statistics that can be found on [NBA.com](
https://www.nba.com/stats/players/shooting/?Season=2020-21&SeasonType=Playoffs&PerMode=Totals). 
It consists of shots taken by four different players during the 2021 NBA 
Playoffs.

----------------------------------------------------------------------

### Data Dictionary

|Variable |Class     |Description                                                  |       
|:--------|:---------|:------------------------------------------------------------|     
|SHOOTER  |string    |Name of the player taking the shot                           |     
|X        |float     |Horizontal distance of the shot taken from the baseline in feet|     
|Y        |float     |Vertical distance of the shot taken from the baseline in feet  |     
|RANGE    |string    |Radius range of the shot taken from the baseline in feet       |     
|DEFENDER |string    |Name of the player defending the shot                        |     
|SCORE    |string    |'MADE' if shot is scored, else 'MISSED'                      |     

----------------------------------------------------------------------     

### Loading Data 

```{r}
#| label: load-data
#| echo: true
library(vijaynbastats)
library(dplyr)
library(here)
library(ggplot2)

nba_players_shooting <- read.csv(here::here("data/nba_players_shooting.csv"))
shots <- nba_players_shooting # shortened version and will be augmented
```

There are `r nrow(shots)` rows and `r ncol(shots)` 
columns (`r get_and_or_list(names(shots))`) in the dataset.

### Statistics To Be Used

#### Some Definitions

| Initialism | Definition                                               |
|:-----------|:---------------------------------------------------------|
| _PTS_        | Points scored                                            |
| _FGA_        | Field goals attempted                                    |
| _FTA_        | Free throws attempted                                    |
| _C~FTA~_     | Free throws attempted multiplier (coeficient)            |
| _TS_         | True shooting percentage                                 |
------------------------------------------------------------------------


    1.  True Shooting Percentage (TS) : A measure of shooting efficiency 
        that takes into account field goals, 3-point field goals, and free 
        throws. It is defined as:   
    
$$
TS = \frac{0.5 * PTS}{FGA + (C_{FTA} * FTA)}
$$

    2.   Average Shooting Defense (ASD): 1 minus the average shooting 
         success of all players against a specific defender.   
    3.   Expected Shooting (ES): The expected shooting success of a
         player against a specific defender. Equal to 1 minus the defender's
         Average Shooting Defense.
    4.   Shooting Success (SS): The shooting success of a player against a
         specific defender. 
    5.   Mean Shooting Success (MSS): The mean of shooting success 
    4.   Relative Shooting Success (RSS): A measure of how efficient a player is
         at scoring points relative to the expected points scored by an average
         player against the same primary defender. Equal to the player's 
         shooting success divided by the defender's Average Shooting Defense.

### Pull In Summary Statistics from Basketball Reference

```{r}
#| label: pull-summary-stats
#| echo: true
#| eval: true
players <- sort(unique(shots$SHOOTER))

summary_stats <- read.csv(here::here(
  "inst/extdata/csv_files/2021_nba_playoff_stats.csv"))
summary_stats <- summary_stats[summary_stats$Player %in% players, ]

```


### Adding Distance and Polar Coordinates for Each Shot

```{r}
#| label: add-distance
#| echo: true
#| eval: true
#| tbl_cap: "Data with Distance Added"

shots <- shots |>
  mutate(DISTANCE = sqrt(X^2 + Y^2),
         POLAR_ANGLE = atan2(Y, X) * 180 / pi)
points <- integer(length(shots$DISTANCE))
three_point <- logical(length(shots$DISTANCE))
for (i in seq_along(shots$DISTANCE)) {
  points[i] <- get_points(shots$X[i], 
                          shots$Y[i], 
                          shots$SCORE[i])
  three_point[i] <- is_three_point(shots$X[i], 
                                   shots$Y[i])
}
shots$POINTS <- points
shots$THREE_POINT <- three_point


sc_threes <- shots[shots$SHOOTER == "Seth Curry", "THREE_POINT"]
sc_points <- shots[shots$SHOOTER == "Seth Curry", "POINTS"]
```

Note that my calculation of what is a three point shot is based on the NBA
rules at https://official.nba.com/rule-no-1-court-dimensions-equipment/ and
the X and Y coordinates of the shots taken. See the function `is_three_point`
for the implementation. However, the official NBA statistics for these games 
has Seth Curry with 
`r summary_stats$X3PA[summary_stats$Player == "Seth Curry"]` 
three point attempts and 
`r summary_stats$X3P[summary_stats$Player == "Seth Curry"]` 
three point shots made. While I have calculated 
`r sum(sc_threes)` three point attempts and 
`r sum(ifelse(sc_points == 3, 1, 0))` three point shots made. The discrepancy
is made up in the two point shots which are missing the 
`r sum(sc_threes) - summary_stats$X3PA[summary_stats$Player == "Seth Curry"]`
shots.

```{r}
#| label: fig-curry-shot-locations
#| echo: true
#| eval: true
# Change the point size, and shape
ggplot(shots[shots$SHOOTER == "Seth Curry", ], aes(x = X, y = Y)) +
  geom_point(size = 2, col = shots[shots$SHOOTER == "Seth Curry", ]$POINTS + 1)
```

### Add Shot Value in Points Based on Shot Distance and Location

```{r}
#| label: add-shot-value
#| echo: true
#| eval: true
#| tbl_cap: "Data with Shot Value Added"

```{r}
#| label: tbl-player-ts
#| echo: true
#| eval: true
#| tbl_cap: "True Shooting Percentage for Each Player During 2021 Playoffs"
players <- sort(unique(shots$SHOOTER))
ts <- numeric(length(players))
for (i in seq_along(players)) {
  player_data <- shots[shots$SHOOTER == players[i], ]
  stats <- summary_stats[summary_stats$Player == players[i], ]
  fta <- stats$FTA
  fga <- stats$FGA
  pts <- stats$PTS
  pts2 <- stats$`2P`
  pts3 <- stats$`3P`
  ft <- stats$FT
  fta <- stats$FTA
  ts[i] <- true_shooting_percentage(pts = pts, 
                                    fga = fga,
                                    fta = fta)
}
ts_tbl <- data.frame(PLAYER = players, TS = ts, stringsAsFactors = FALSE)
ts_tbl |>
  dplyr::arrange(desc(TS)) |>
  knitr::kable()
```


